---
title: "Edinburgh bin sensor project - Shiny app"
output:
  html_document:
    df_print: paged
runtime: shiny
---

# Introduction

This document creates a small Shiny app to compare the rubbish weight between 2 streets which had bins with bin sensors. 

## Inital data preparation

Begin with some initial data preparation:
```{r, message=FALSE, warning=FALSE}
library(shiny)
library(tidyverse)
library(sf)

#read in the cleaned and combined bin & osm data 
combined_bin_osm <- st_read(here::here("cleaned_data/combined_bin_osm_data/combined_bin_osm.shp"), quiet = TRUE)

  
#undoing some shortened naming when read to shape file
combined_bin_osm <- combined_bin_osm %>%
mutate(cumul_date = date - min(combined_bin_osm$date)) %>%
  rename("street_name" = "strt_nm",
        "total_vol_l3" = "ttl_v_3",
       "total_weight_kg" = "ttl_wg_",
       "cumul_total_vol_l3" = "cm_t__3",
       "cumul_total_weight_kg" = "cm_tt__",
       "highway_group" = "hghwy_g")

# read in all the .shp files in the cleaned osm data folder 
root_of_project_path <- here::here()
project_path <- '/cleaned_data/osm_data/'
dir_path <- paste0(root_of_project_path, project_path)
  
file_pattern <- '*.shp'
shp_files <- list.files(dir_path, pattern = file_pattern)

for (i in seq_along(shp_files)) {
  assign(str_remove(shp_files[i], ".shp"), st_read(paste0(dir_path, shp_files[i]), quiet = TRUE))
}

#undoing some shortened naming when read to shape file
streets_simplified <- streets_simplified %>%
  rename("street_name" = "strt_nm",
         "highway_group" = "hghwy_g")

combined_bin_osm <- combined_bin_osm %>%
    mutate(street_name = str_to_title(street_name))

min_max_coords <- c(ymin= 55.928479867725436,
                    xmin= -3.227624857214751,
                    ymax= 55.98333902701634,
                    xmax= -3.140735628435386)

blankbg <-theme(axis.line=element_blank(),
                axis.text.x=element_blank(),
                axis.text.y=element_blank(),
                axis.ticks=element_blank(),
                axis.title.x=element_blank(), 
                axis.title.y=element_blank(),
                legend.position = "none",
                plot.background=element_blank(),
                panel.grid.minor=element_blank(),
                panel.background=element_blank(),
                panel.grid.major=element_blank(),

)


```

## Shiny app

And then the Shiny app. The user can select 2 streets and compare the total weight, how the weight accumulated over time and where the streets are located spatially. The user can select for the graphs to be in kilograms or equivalent weight object (as an alternative way of conceptualising the weight of the rubbish).  

```{r}
shinyApp(
  
  ui = fluidPage(
    fluidRow(column(3, selectInput("street1_chosen", "Street 1:",
                choices = unique(combined_bin_osm$street_name),
                selected = "Atholl Crescent")),
    column(3, selectInput("street2_chosen", "Street 2:",
                choices = unique(combined_bin_osm$street_name),
                selected = "Calton Hill")),
    column(3, sliderInput("date_chosen", "Days:", 
                          min(unique(combined_bin_osm$cumul_date)),  
                          max(unique(combined_bin_osm$cumul_date)),
                          max(unique(combined_bin_osm$cumul_date)), 
                          animate = animationOptions(interval = 1400))),
    column(3, selectInput("scale_chosen", "Scale choice:",
                              choices = c("Kilograms", "Objects")))),
    fluidRow(column(6, plotOutput("map")),
    column(6, plotOutput("street_plot"))),
    fluidRow(plotOutput("line_plot"))
  ),

  server = function(input, output) {

    
    subset_bin_data <- reactive({
    
     streets_chosen <- c(input$street1_chosen, input$street2_chosen)
      
      combined_bin_osm %>%
        filter(cumul_date == input$date_chosen, street_name %in% streets_chosen) 
      
    })


     output$map = renderPlot({

ggplot() +
         blankbg +
  geom_sf(data = water,
          fill = "steelblue",
          # size = .8,
          lwd = 0,
          alpha = .3) +
  geom_sf(data = park_multipoly,
          fill = "green",
          # size = .8,
          lwd = 0,
          alpha = .3) +
  geom_sf(data = park_poly,
          fill = "green",
          # size = .8,
          lwd = 0,
          alpha = .3) +
  geom_sf(data = railways,
          color = "grey30",
          size = .2,
          linetype="dotdash",
          alpha = .5) +
  geom_sf(data = streets_simplified %>% 
            filter(highway_group == "small"),
          size = .1,
          color = "grey40") +
  geom_sf(data = streets_simplified %>% 
            filter(highway_group == "medium"),
          size = .3,
          color = "grey35") +
  geom_sf(data = streets_simplified %>% 
            filter(highway_group == "large"),
          size = .5,
          color = "grey30") +
  geom_sf(data =   subset_bin_data(), aes(color = street_name), size = 1.5, show.legend = "line") +
  scale_colour_manual(values = c("#e41a1c", "#377eb8")) +
  labs(caption = 'Edinburgh', size = 2) +
  coord_sf(ylim = c(min_max_coords[1], min_max_coords[3]),
           xlim = c(min_max_coords[2], min_max_coords[4]),
           expand = FALSE)
     })

    
    output$street_plot = renderPlot({
     
      p <- subset_bin_data() %>%
        ggplot(aes(x = street_name, y = cumul_total_weight_kg, fill = street_name)) +
        geom_col() +
        scale_fill_manual(values = c("#e41a1c", "#377eb8")) +
        theme_minimal() 
      
      if (input$scale_chosen == "Kilograms") {
        p + 
        scale_y_continuous(labels = scales::comma) +
        labs(x = "Street name", y = "Cumulative total rubbish weight (kg)", fill = "Street name")
          
      } else {
        p +
        scale_y_continuous(breaks = c(3, 120, 980, 4300, 12000, 19500, 30000) , labels = c('Brick', 'Baby elephant', 'Vauxhall Corsa', 'Adult Elephant', 'Bus', 'Fire engine', 'Humpback whale')) +
        labs(x = "Street name", y = "Cumulative total rubbish weight\n(equivalent weight objects)", fill = "Street name")
      }
    })
  
    
    output$line_plot = renderPlot({
    
      streets_chosen <- c(input$street1_chosen, input$street2_chosen)
      
      p <- combined_bin_osm %>%
        filter(cumul_date <= input$date_chosen, street_name %in% streets_chosen) %>%
        ggplot(aes(x = date, y = cumul_total_weight_kg, color = street_name)) +
        geom_line() +
        scale_colour_manual(values = c("#e41a1c", "#377eb8"))  +
        xlim(min(combined_bin_osm$date), max(combined_bin_osm$date)) +
        scale_x_date(date_labels = "%d%b%y") +
        theme_minimal() 
      
       if (input$scale_chosen == "Kilograms") {
        p +
        scale_y_continuous(labels = scales::comma) +
        labs(x = "Street name", y = "Cumulative total rubbish weight (kg)", color = "Street name")
       } else {
        p +
        scale_y_continuous(breaks = c(3, 120, 980, 4300, 12000, 19500, 30000) , labels = c('Brick', 'Baby elephant', 'Vauxhall Corsa', 'Adult Elephant', 'Bus', 'Fire engine', 'Humpback whale')) +
        labs(x = "Street name", y = "Cumulative total rubbish weight\n(equivalent weight objects)", color = "Street name")
       }
    
    })
  },

  options = list(height = 500)
)
```

